Further Evidence of Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-01-20 src/ms.tex
paper edits

\textbf{We matched 10844 LAMOST targets to 10650 unique Kepler IDs in the \citet{McQuillan2014} sample. In the \citet{Santos2021} sample we found 54982 unique cross-matched sources in LAMOST, of which 18990 have published temperatures and rotation periods. For the Kepler sources with duplicate cross-matched LAMOST sources we kept the source with a brighter Gaia DR2 $G$ magnitude. We additionally selected sources with 3000~K~$<~\teff~<$~8000~K, 3~$<$~\logg~$<$~5~dex, and -2~$<$~[Fe/H]~$<$~2~dex. The final sample contains 10516 unique targets with well-determined \prot and \teff (having a median error of 24~K). The metallicity distribution of the LAMOST--McQuillan sample is centered near solar, with a mean and standard deviation of -0.1~dex and 0.26~dex, respectively. There is negligible overlap (3 stars) between our LAMOST--McQuillan sample and the CKS sample since the \citet{McQuillan2014} did not publish rotation periods for KOIs, which were the targets of the CKS project. Rotation periods for KOIs are instead published in \citet{McQuillan2013}, as discussed in \S\ref{subsec:cks}. Visualizations of the \lamostmcq and \lamostsan samples are shown in Figure~\ref{fig:xmatch}.}

@@

\begin{deluxetable*}{lllll}
\tabletypesize{\scriptsize}
\tablecolumns{5}
\tablewidth{0pt}
\tablecaption{Results of constant Rossby model fits.\label{tab:mcmc}}
\tablehead{
\colhead{Sample} & 
\colhead{\teff range} & 
\colhead{Ro} & 
\colhead{f} & 
\colhead{$T_\mathrm{sys}$ (K)} 
}
\startdata
\textit{Short-period pile-up} \\
\hline
\lamostmcq 10th \prot pctl. & 4500~K--5800~K & 0.367 $\pm$ 0.009 & 0.008 $\pm$ 0.006 &  44 $\pm$ 36 \\
\hline 
\textit{Long-period pile-up} \\
\hline
\lamostmcq 90th \prot pctl. & 5000~K--6250~K & 1.38 $\pm$ 0.04 & 0.011 $\pm$ 0.008 & 49 $\pm$ 19 \\
\lamostmcq 90th \prot pctl. & 5800~K--6250~K & 1.4 $\pm$ 0.1 & 0.019 $\pm$ 0.015 & 44 $\pm$ 43 \\
CKS (CKS \teff) & 5800~K--6250~K & 1.77 $\pm$ 0.07 & 0.02 $\pm$ 0.01 & -70 $\pm$ 15 \\ 
CKS (SPOCS \teff) & 5800~K--6250~K & 1.9 $\pm$ 0.1 & 0.12 $\pm$ 0.01 & -136 $\pm$ 23 \\
CKS (M19 \teff) & 5800~K--6250~K & 1.6 $\pm$ 0.2 & 0.18 $\pm$ 0.02 & 21 $\pm$ 42 \\
\citet{Hall2021} main-sequence & 5000~K--6250~K & 1.7 $\pm$ 0.2 & 0.30 $\pm$ 0.04 & -5 $\pm$ 64 \\
\citet{Hall2021} main-sequence & 5800~K--6250~K & 1.9 $\pm$ 0.4 & 0.27 $\pm$ 0.05 & -29 $\pm$ 79 \\
%All samples & 5000~K--6250~K & 1.47 $\pm$ 0.05 & 0.132 $\pm$ 0.008 & 2 $\pm$ 23 (LAMOST) \\
% & & & & 10 $\pm$ 15 (CKS) \\
% & & & & 61 $\pm$ 21  (Hall) \\
%All samples & 5800~K--6250~K & 1.67 $\pm$ 0.08 & 0.116 $\pm$ 0.009 & -42 $\pm$ 24 (LAMOST) \\
%&  &  &  & -44 $\pm$ 20 (CKS) \\
%&  &  &  & 3 $\pm$ 23 (Hall) \\ 
\enddata
%\tablecomments{}
\end{deluxetable*}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/fraction.pdf}
    \caption{The relative fraction of stars on the long- and short-period pile-ups, in the \lamostmcq sample, in 10~K windows of \teff.}
    \label{fig:fraction}
\end{figure}

@@

    \item An increasing number of open clusters are being discovered by searching for a clustering of rotation periods along a slow-rotator sequence in the \teff–\prot or color–\prot plane. However, the long-period pile up discovered here can mimic a slow-rotator sequence in a small sample of unassociated stars with different ages and precisely measured temperatures and rotation periods. This is clearly demonstrated by the CKS sample in the right-hand panel of Figure~\ref{fig:surveys}. Taken out of context, this sample resembles a group of coeval stars with a slow-rotator sequence. The discovery of pile-ups in stellar rotation periods therefore has consequences for open cluster studies. When an overdensity or ridge is present in the rotation period distribution of a stellar population, care must be taken to ensure that it is not caused by WMB or core-envelope coupling before assuming that population is coeval and using the overdensity to age-date it.

>>>>

\textbf{We matched 10844 LAMOST targets to 10550 unique Kepler IDs in the \citet{McQuillan2014} sample. In the \citet{Santos2021} sample we found 54982 unique cross-matched sources in LAMOST, of which 18990 have published temperatures and rotation periods. For the Kepler sources with duplicate cross-matched LAMOST sources we kept the source with a brighter Gaia DR2 $G$ magnitude. We additionally selected sources with 3000~K~$<~\teff~<$~8000~K, 3~$<$~\logg~$<$~5~dex, and -2~$<$~[Fe/H]~$<$~2~dex. The final sample contains 10516 unique targets with well-determined \prot and \teff (having a median error of 24~K). The metallicity distribution of the LAMOST--McQuillan sample is centered near solar, with a mean and standard deviation of -0.1~dex and 0.26~dex, respectively. There is negligible overlap (3 stars) between our LAMOST--McQuillan sample and the CKS sample since the \citet{McQuillan2014} did not publish rotation periods for KOIs, which were the targets of the CKS project. Rotation periods for KOIs are instead published in \citet{McQuillan2013}, as discussed in \S\ref{subsec:cks}. Visualizations of the \lamostmcq and \lamostsan samples are shown in Figure~\ref{fig:xmatch}.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{./figures/rocrit.pdf}
    \caption{}
    \label{fig:rocrit}
\end{figure*}

\begin{deluxetable*}{lllll}
\tabletypesize{\scriptsize}
\tablecolumns{5}
\tablewidth{0pt}
\tablecaption{Results of constant Rossby model fits.\label{tab:mcmc}}
\tablehead{
\colhead{Sample} & 
\colhead{\teff range} & 
\colhead{Ro} & 
\colhead{f} & 
\colhead{$T_\mathrm{sys}$ (K)} 
}
\startdata
\textit{Short-period pile-up} \\
\hline
\lamostmcq 10th \prot pctl. & 4500~K--5800~K & 0.37 $\pm$ 0.02 & 0.011 $\pm$ 0.009 &  46 $\pm$ 80 \\
\hline 
\textit{Long-period pile-up} \\
\hline
\lamostmcq 90th \prot pctl. & 5000~K--6250~K & 1.38 $\pm$ 0.04 & 0.011 $\pm$ 0.008 & 49 $\pm$ 19 \\
\lamostmcq 90th \prot pctl. & 5800~K--6250~K & 1.4 $\pm$ 0.1 & 0.019 $\pm$ 0.015 & 44 $\pm$ 43 \\
CKS (CKS \teff) & 5800~K--6250~K & 1.77 $\pm$ 0.07 & 0.02 $\pm$ 0.01 & -70 $\pm$ 15 \\ 
CKS (SPOCS \teff) & 5800~K--6250~K & 1.9 $\pm$ 0.1 & 0.12 $\pm$ 0.01 & -136 $\pm$ 23 \\
CKS (M19 \teff) & 5800~K--6250~K & 1.6 $\pm$ 0.2 & 0.18 $\pm$ 0.02 & 21 $\pm$ 42 \\
\citet{Hall2021} main-sequence & 5000~K--6250~K & 1.7 $\pm$ 0.2 & 0.30 $\pm$ 0.04 & -5 $\pm$ 64 \\
\citet{Hall2021} main-sequence & 5800~K--6250~K & 1.9 $\pm$ 0.4 & 0.27 $\pm$ 0.05 & -29 $\pm$ 79 \\
%All samples & 5000~K--6250~K & 1.47 $\pm$ 0.05 & 0.132 $\pm$ 0.008 & 2 $\pm$ 23 (LAMOST) \\
% & & & & 10 $\pm$ 15 (CKS) \\
% & & & & 61 $\pm$ 21  (Hall) \\
%All samples & 5800~K--6250~K & 1.67 $\pm$ 0.08 & 0.116 $\pm$ 0.009 & -42 $\pm$ 24 (LAMOST) \\
%&  &  &  & -44 $\pm$ 20 (CKS) \\
%&  &  &  & 3 $\pm$ 23 (Hall) \\ 
\enddata
%\tablecomments{}
\end{deluxetable*}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/fraction.pdf}
    \caption{The relative fraction of stars on the long- and short-period pile-ups, in the \lamostmcq sample, in 10~K windows of \teff (for stars with \logg~$>$4.1 dex).}
    \label{fig:fraction}
\end{figure}

@@

    \item {\color{red} An increasing number of open clusters are being discovered by searching for a clustering of rotation periods along a slow-rotator sequence in the \teff–\prot or color–\prot plane. However, the long-period pile up discovered here can mimic a slow-rotator sequence in a small sample of unassociated stars with different ages and precisely measured temperatures and rotation periods. This is clearly demonstrated by the CKS sample in the right-hand panel of Figure~\ref{fig:surveys}. Taken out of context, this sample resembles a group of coeval stars with a slow-rotator sequence. The discovery of pile-ups in stellar rotation periods therefore has consequences for open cluster studies. When an overdensity or ridge is present in the rotation period distribution of a stellar population, care must be taken to ensure that it is not caused by WMB or core-envelope coupling before assuming that population is coeval and using the overdensity to age-date it.}
